Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands

### 2024-10-31 index.qmd
Save

Dryland regions, encompassing arid, semi-arid, hyper-arid, and dry sub-humid systems, account for 40% of the Earth’s land surface. These regions are home to approximately 2 billion people globally and constitute the largest terrestrial biome (**iucn-drylands-2019)**. Despite supporting diverse ecosystems and human populations, dryland regions face mounting challenges exacerbated by increasing urbanization, expanding agricultural activities, and climate-induced amplification of precipitation patterns [@taylor2013]. This water scarcity is intensifying due to the compounding effects of climate variability and increased groundwater extraction [@taylor2013]. In drylands, groundwater serves as a vital resource for sustaining ecosystems and meeting human needs [@scanlon2006; @yao2018].

Base flow is the sustained portion of streamflow in the absence of runoff that is derived from groundwater discharge **(usgs-glossary-2018).** Base flow is critical to maintaining seasonal low-flow regimes, supporting aquatic ecosystems, and facilitating the transport of nutrients and chemicals. Base-flow contribution to streamflow can be highly variable spatially [@singh2018; @bosch2017; @beck2013], and temporally [@ficklin2016; @tan2020]. Increasing groundwater extraction, changes in land cover/land use, and changes in precipitation patterns due to climate change affect the timing and volumes of base flow [@tan2020; @taylor2013]. Effective management of water quantity and quality depends on understanding seasonal and interannual base flow patterns and long-term changes in base flow behavior.

@@

This study develops a technique for estimating BFI in ungauged basins in Arizona and determines the state’s long-term mean BFI from 1991 to 2020. Due to interannual variability in BFI, a 30-year period of record is used to evaluate the characteristic base-flow dynamics of Arizona. Using a machine learning model trained with the hydrogeologic characteristics of surface water basins, we first determine annual BFIs for each year in the 1991-2020 study period for ungauged river reaches and physiographic regions of Arizona. Long-term mean BFI data for the study period are then estimated. Regional patterns and trends in base flow of instrumented sites are subsequently investigated. These results provide novel analyses of low flow processes in Arizona...

@@

Arizona's hydrology varies seasonally and spatially between its physiographic regions. In the summer, localized and intense convective precipitation events are driven by the North American Monsoon, while in the winter, orographic precipitation comes from Pacific frontal systems **(eastoe2019mtnblock).** While monsoonal precipitation can account for up to 50% of annual precipitation, evaporation and dry preceding soil properties leads to most precipitation becoming runoff [@sheppard2002]. As such, 94% of streams in Arizona are ephemeral or intermittent **(epa_AZephemeral).** Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack. While winter precipitation provides only 30% of annual averages, it provides the bulk of water for natural reservoirs [@sheppard2002].

@@

Daily observed streamflow data obtained from the United States Geological Survey (USGS) National Water Information System (NWIS) were used in this study. Streamgages were selected depending on criteria to ensure the applicability of each site. @odonnell2016 found that 8-10 years of calibration data were necessary to account for climate variability in paired watershed studies in the region. Following this criteria, a minimum of 10 years of record was chosen. Any years missing more than 30 days of streamflow were removed from the possible record period. This analysis focuses on natural, streamflow-influencing recharge, excluding streamgages affected by regulation or diversions. These gages were identified through annual reports on water data published by the USGS **(usgs2010)**. Finally, streamgages on the Colorado River were excluded from the dataset on the basis that they represent managed flows shared by Colorado River Compact states. Following these parameters, 206 USGS streamgages with acceptable periods of record were used in this study (Figure \###### study-area).

@@

The annual precipitation and temperature data came from the PRISM climate group at Oregon State University at a resolution of 4 km (<https://prism.oregonstate.edu;> [@daly2008]). The PRISM dataset provides valuable insights into regional climate in ungauged regions and has been shown to perform well across the southwestern US **(buban_PRISM).** Instead of the water year, PRISM data uses a calendar-year format, which was adopted for consistency in the water balance. Although this may introduce challenges in the annual estimates due to inter-annual snow storage, the use of long-term annual averages is likely to reduce any potential errors [@reitz2017].

@@

where $b$ is base flow, $\alpha$ is the filter parameter, $Q$ is the total streamflow, and $k$ is the time step. A filter parameter $\alpha$ of 0.925 was used as in @nathan1990 and **\@fuka2014ecohydrology**. The filter was run three times (forward, backward, forward) to attenuate the base-flow signal. @eckhardt2023 noted that recursive digital filters lack a physical basis, but as the method is easy to automate, objective, and repeatable, it is appropriate for a regional-scale study.

@@

The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018] **(schmidt_ML_2020, Rozos2021Machine)**. In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm **(chen2016xgboost)** to predict BFI at ungauged locations using catchment characteristics as predictors. The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems. It has grown in popularity in recent years due to its rapid, efficient, and scalable nature, and has proven effective in similar environmental modeling tasks such as streamflow forecasting **(szczepanek2022daily, ni2020streamflow)** and classifying land use/land cover data **(Georganos2018Very)**.

XGBoost uses gradient boosting on decision tree algorithms to leverage the combination of multiple very low variance models into one overall prediction. Gradient boosting works sequentially, with the first tree being learned on the known target values, and each subsequent tree being learned on the errors of the tree prior. The algorithm then applies a weighting to each tree based on this boosting to determine how much the tree will contribute towards the final output. The final prediction is obtained by the input data traversing all $n$ weighted trees in the learned ensemble. The trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (Table \ref{tab-######}. Certain predictors were subdivided according to their areal coverage of each basin (e.g. land cover was divided into 16 subdivisions).

>>>>

Dryland regions, encompassing arid, semi-arid, hyper-arid, and dry sub-humid systems, account for 40% of the Earth’s land surface. These regions are home to approximately 2 billion people globally and constitute the largest terrestrial biome [@iucn-drylands-2019]. Despite supporting diverse ecosystems and human populations, dryland regions face mounting challenges exacerbated by increasing urbanization, expanding agricultural activities, and climate-induced amplification of precipitation patterns [@taylor2013]. This water scarcity is intensifying due to the compounding effects of climate variability and increased groundwater extraction [@taylor2013]. In drylands, groundwater serves as a vital resource for sustaining ecosystems and meeting human needs [@scanlon2006; @yao2018].

Base flow is the sustained portion of streamflow in the absence of runoff that is derived from groundwater discharge [@usgs-glossary-2018]. Base flow is critical to maintaining seasonal low-flow regimes, supporting aquatic ecosystems, and facilitating the transport of nutrients and chemicals. Base-flow contribution to streamflow can be highly variable spatially [@singh2018; @bosch2017; @beck2013], and temporally [@ficklin2016; @tan2020]. Increasing groundwater extraction, changes in land cover/land use, and changes in precipitation patterns due to climate change affect the timing and volumes of base flow [@tan2020; @taylor2013]. Effective management of water quantity and quality depends on understanding seasonal and interannual base flow patterns and long-term changes in base flow behavior.

@@

This study develops a technique for estimating BFI in ungauged basins in Arizona and determines the state’s long-term mean BFI from 1991 to 2020. Due to interannual variability in BFI, a 30-year period of record is used to evaluate the characteristic base-flow dynamics of Arizona. Using a machine learning model trained with the hydrogeologic characteristics of surface water basins, we first determine annual BFIs for each year in the 1991-2020 study period for ungauged river reaches and physiographic regions of Arizona. Long-term mean BFI data for the study period are then estimated. Regional patterns and trends in base flow of instrumented sites are subsequently investigated. These results provide novel analyses of low flow processes in Arizona...######

@@

Arizona's hydrology varies seasonally and spatially between its physiographic regions. In the summer, localized and intense convective precipitation events are driven by the North American Monsoon, while in the winter, orographic precipitation comes from Pacific frontal systems [@eastoe2019mtnblock]. While monsoonal precipitation can account for up to 50% of annual precipitation, evaporation and dry preceding soil properties leads to most precipitation becoming runoff [@sheppard2002]. As such, 94% of streams in Arizona are ephemeral or intermittent [@epa_AZephemeral]. Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack. While winter precipitation provides only 30% of annual averages, it provides the bulk of water for natural reservoirs [@sheppard2002].

@@

Daily observed streamflow data obtained from the United States Geological Survey (USGS) National Water Information System (NWIS) were used in this study. Streamgages were selected depending on criteria to ensure the applicability of each site. @odonnell2016 found that 8-10 years of calibration data were necessary to account for climate variability in paired watershed studies in the region. Following this criteria, a minimum of 10 years of record was chosen. Any years missing more than 30 days of streamflow were removed from the possible record period. This analysis focuses on natural, streamflow-influencing recharge, excluding streamgages affected by regulation or diversions. These gages were identified through annual reports on water data published by the USGS [@usgs2010]. Finally, streamgages on the Colorado River were excluded from the dataset on the basis that they represent managed flows shared by Colorado River Compact states. Following these parameters, 206 USGS streamgages with acceptable periods of record were used in this study (Figure \###### study-area).

@@

The annual precipitation and temperature data came from the PRISM climate group at Oregon State University at a resolution of 4 km (<https://prism.oregonstate.edu;> [@daly2008]). The PRISM dataset provides valuable insights into regional climate in ungauged regions and has been shown to perform well across the southwestern US [@buban_PRISM]**.** Instead of the water year, PRISM data uses a calendar-year format, which was adopted for consistency in the water balance. Although this may introduce challenges in the annual estimates due to inter-annual snow storage, the use of long-term annual averages is likely to reduce any potential errors [@reitz2017].

@@

where $b$ is base flow, $\alpha$ is the filter parameter, $Q$ is the total streamflow, and $k$ is the time step. A filter parameter $\alpha$ of 0.925 was used as in @nathan1990 and @fuka2014ecohydrology. The filter was run three times (forward, backward, forward) to attenuate the base-flow signal. @eckhardt2023 noted that recursive digital filters lack a physical basis, but as the method is easy to automate, objective, and repeatable, it is appropriate for a regional-scale study.

@@

The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018; @schmidt_ML_2020; @Rozos2021Machine]. In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm **(chen2016xgboost)** to predict BFI at ungauged locations using catchment characteristics as predictors. The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems. It has grown in popularity in recent years due to its rapid, efficient, and scalable nature, and has proven effective in similar environmental modeling tasks such as streamflow forecasting [@szczepanek2022daily**; @ni2020streamflow**] and classifying land use/land cover data [@Georganos2018Very].

XGBoost uses gradient boosting on decision tree algorithms to leverage the combination of multiple very low variance models into one overall prediction. Gradient boosting works sequentially, with the first tree being learned on the known target values, and each subsequent tree being learned on the errors of the tree prior. The algorithm then applies a weighting to each tree based on this boosting to determine how much the tree will contribute towards the final output. The final prediction is obtained by the input data traversing all $n$ weighted trees in the learned ensemble. The trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (Table \ref{tab-######}##}##}##}. Certain predictors were subdivided according to their areal coverage of each basin (e.g. land cover was divided into 16 subdivisions).
